Generalized Tractability

### 2023-06-06 GeneralizedTractability_230529.tex
Updates from Overleaf

\section{Tractability on a Restricted Domain}

Suppose that the domain of interest $(\varepsilon^{-1},d)$ is not all of $(0,\infty) \times \naturals$ but some subset
\begin{equation}
    \Omega  := \{(\varepsilon^{-1},d) : \varepsilon > E(d), \ d \in \naturals \}, \qquad \text{where }E : \naturals \to [0,\infty).
\end{equation}
We then expect that the equivalent conditions for (strong) tractability will be similar in form, but weaker than, the conditions in Theorems  \ref{thm_main_strong_tract2} and \ref{thm_main_tract2}.

@@

Define $M: \naturals \to \natzero \cup \{\infty\}$
\begin{equation} \label{eq:Mddef}
    M(d) := \begin{cases}
        \min\{n \in \natzero : \lambda_{n+1,d} \le E(d)\}, & E(d) > 0, \\
        \infty, & E(d) = 0.
    \end{cases}
\end{equation}
This means that 
\[
\left \{
\begin{array}{lll}
&(\lambda_{1,d}^{-1},d), (\lambda_{2,d}^{-1},d), \ldots \notin \Omega, & M(d) = 0, \\
\lambda_{1,d}\ge  \cdots \ge \lambda_{M(d),d} > E(d) \ge\lambda_{M(d)+1,d}, &
(\lambda_{1,d}^{-1},d), \ldots, (\lambda^{-1}_{M(d),d},d)  \in \Omega, & 0 < M(d) < \infty, \\
&
(\lambda_{1,d}^{-1},d), (\lambda_{2,d}^{-1},d), \ldots \in \Omega, & M(d) = \infty.
\end{array}
\right .
\]
Also, for each $d \in \naturals$ define the interval 
\begin{equation} \label{eq:hIdef}
\hI_{d}:=\begin{cases} 
\emptyset, & M(d) = 0, \\
\{1,\ldots, M(d)\}, & 0 < M(d) < \infty, \\
\naturals, & M(d) = \infty.
\end{cases}
\end{equation}
From the definition of $M(d)$ it follows that $(\lambda_{i,d}^{-1},d) \in \Omega  \iff i \in \hI_d$, and
\[
\comp(\varepsilon,d) \in \hI_d \qquad \forall (\varepsilon^{-1},d) \in \Omega.
\]

@@

\begin{theorem}\label{thm_main_rest_strong_tract}
Let $T$ be a tractability function as specified in \eqref{eq:Tspec} and satisfying \eqref{eq:Tconditions} and \eqref{eq:ptauassume}.  A problem is strongly $T$-tractable \emph{on the restricted domain $\Omega$} iff there exists $\vp \in [\vzero, \boldsymbol{\infty})$ and an integer $L_{\vp} > 0$ such that
\begin{equation} \label{eq:strong_rest_tractiff}
     S_{\vp}:=\sup_{d \in \naturals} \sum_{i = L_{\vp}}^{M(d)} \frac{1}{T(\lambda_{i,d}^{-1},1,\vp)} < \infty.
\end{equation}
By convention, if $M(d) = 0$, the sum is zero.

@@

\begin{proof}
\textbf{Sufficient condition:}\newline
Fix $d$.  For the case where $M(d)> 0$, following the argument for the sufficient condition in Theorem \ref{thm_main_strong_tract2} it follows that for all $d' \in \naturals$ and integers $n$ and $N$ with $1 \le N \le n \le M(d)$, we have
\begin{align*}
    \lambda_{n+1,d} \le \lambda_{n,d} \le \cdots \le \lambda_{N,d}
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},d',\vp)} \le \frac{1}{T(\lambda_{n, d}^{-1},d',\vp)} \le \cdots \le \frac{1}{T(\lambda_{N, d}^{-1},d',\vp)} \\
     \implies \frac{1}{T(\lambda_{n+1, d}^{-1},d',\vp) }
    \le &\frac{1}{n-N+1} \sum_{i=N}^n  \frac{1}{T(\lambda_{i, d}^{-1},d',\vp) }
    \le \frac{1}{n-N+1} \sum_{i=N}^{M(d)}  \frac{1}{T(\lambda_{i, d}^{-1},d',\vp)}.
\end{align*}
Thus, we can conclude from the previous line that
\begin{multline} \label{eq:implicate_a_rest}
    n - N +1 \ge T(\varepsilon^{-1},d',\vp) \sum_{i=N}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},d',\vp)} \\
   \implies   \frac{1}{T(\lambda_{n+1, d}^{-1},d',\vp)} \le
   \frac{1}{n-N+1} \sum_{i=N}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},d',\vp) } \le \frac{1}{T(\varepsilon^{-1},d',\vp)} \\ \forall d', n, N \in \naturals, \ N \le n \le M(d).
\end{multline}

@@

>>>>

\section{(Strong) Tractability on a Restricted Domain}

Suppose that the domain of interest $(\varepsilon^{-1},d)$ is not all of $(0,\infty) \times \naturals$ but some subset
\begin{equation}
    \Omega  := \{(\varepsilon^{-1},d) : (E(d),\infty), \ d \in \naturals \}, \qquad \text{where }E : \naturals \to [0,\infty).
\end{equation}
We then expect that the equivalent conditions for (strong) tractability will be similar in form, but weaker than, the conditions in Theorems  \ref{thm_main_strong_tract2} and \ref{thm_main_tract2}.

@@

Define $M: \naturals \to \natzero \cup \{\infty\}$
\begin{equation} \label{eq:Mddef}
    M(d) := \begin{cases}
        \min\{n \in \natzero : \lambda_{n+1,d} \le E(d)\}, & E(d) > 0, \\
        \infty, & E(d) = 0.
    \end{cases}
\end{equation}
Also, for each $d \in \naturals$ define the interval 
\begin{equation} \label{eq:hIdef}
\hI_{d}:=\begin{cases} 
\emptyset, & M(d) = 0, \\
\{1,\ldots, M(d)\}, & 0 < M(d) < \infty, \\
\naturals, & M(d) = \infty.
\end{cases}
\end{equation}
From the definition of $M(d)$ it follows that $(\lambda_{i,d}^{-1},d) \in \Omega  \iff i \in \hI_d$, and
\begin{equation} \label{eq:comprestr}
  \comp(\varepsilon,d) \begin{cases}
= 0, & M(d) = 0, \\
    \in \hI_d, & M(d) > 0, 
\end{cases}
\qquad \forall (\varepsilon^{-1},d) \in \Omega.
\end{equation}

@@

\begin{theorem}\label{thm_main_rest_strong_tract}
Let $T$ be a tractability function as specified in \eqref{eq:Tspec} and satisfying \eqref{eq:Tconditions} and \eqref{eq:ptauassume}.  A problem is strongly $T$-tractable \emph{on the restricted domain $\Omega$} iff there exists $\vp \in [\vzero, \boldsymbol{\infty})$ and an integer $L_{\vp} > 0$ such that
\begin{equation} \label{eq:strong_rest_tractiff}
     S_{\vp}:=\sup_{d \in \naturals} \sum_{i = L_{\vp}}^{M(d)} \frac{1}{T(\lambda_{i,d}^{-1},1,\vp)} < \infty.
\end{equation}
By convention, if $M(d) < L_{\vp}$, the sum is zero.

@@

\begin{proof}
\textbf{Sufficient condition:}\newline
Fix $d$.  If $M(d) = 0$, then $\comp(\varepsilon,d) = 0 \le C_{\vp}\, T(\varepsilon^{-1},1,\vp)$ automatically for all $\varepsilon \in [E(d), \infty)$.

For the case $M(d)> 0$, following the argument for the sufficient condition in Theorem \ref{thm_main_strong_tract2} it follows that for all $d' \in \naturals$ and integers $n, N \in \hI_d$  with $N \le n$, we have
\begin{align*}
    \lambda_{n+1,d} \le \lambda_{n,d} \le \cdots \le \lambda_{N,d}
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},d',\vp)} \le \frac{1}{T(\lambda_{n, d}^{-1},d',\vp)} \le \cdots \le \frac{1}{T(\lambda_{N, d}^{-1},d',\vp)} \\
     \implies \frac{1}{T(\lambda_{n+1, d}^{-1},d',\vp) }
    \le &\frac{1}{n-N+1} \sum_{i=N}^n  \frac{1}{T(\lambda_{i, d}^{-1},d',\vp) }
    \le \frac{1}{n-N+1} \sum_{i=N}^{M(d)}  \frac{1}{T(\lambda_{i, d}^{-1},d',\vp)}.
\end{align*}
Thus, we can conclude from the previous line that
\begin{multline} \label{eq:implicate_a_rest}
    n - N +1 \ge T(\varepsilon^{-1},d',\vp) \sum_{i=N}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},d',\vp)} \\
   \implies   \frac{1}{T(\lambda_{n+1, d}^{-1},d',\vp)} \le
   \frac{1}{n-N+1} \sum_{i=N}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},d',\vp) } \le \frac{1}{T(\varepsilon^{-1},d',\vp)} \\ \forall d' \in \naturals,\  n, N \in \hI_d,   \ N \le n.
\end{multline}

Moreover, given that $T$ is increasing in its arguments, we have the following equivalent expression for the complexity via \eqref{eq:comp_def}:
\begin{align} %\label{eq:comp_equiv}
	\nonumber
	\comp(\varepsilon, d) & = \min \{n \in \hI_d : \lambda_{n+1, d} \le \varepsilon\} \qquad \text{by \eqref{eq:comprestr}} \\
	\nonumber
	& = \min \biggl\{n \in \hI_d : \frac{1}{T(\lambda_{n+1,d}^{-1},d',\vp)}\le \frac{1}{T(\varepsilon^{-1},d',\vp) }\biggr\} \qquad  \forall d'\in \naturals\\
	\nonumber
	& \le  \min \biggl\{n \in \hI_d : n \ge N - 1 + T(\varepsilon^{-1},d',\vp) \sum_{i=N}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},d',\vp)} \biggr\} \;  \forall  d', \ 1 \le N \le M(d) \\
 \nonumber
 & \hspace{20ex}\text{by \eqref{eq:implicate_a_rest}.} 
 \intertext{
 This last statement holds for all $N \le M(d)$ because $n \ge N$ automatically in the expression above.  Thus,}
 \comp(\varepsilon,d)
	& \le  T(\varepsilon^{-1},d',\vp) \Biggl [\frac{N}{T(\varepsilon^{-1},d',\vp)}  +  \sum_{i=N}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},d',\vp)} \Biggr] \quad \forall  d',\ 1 \le N \le M(d). \label{eq:compUP}
\end{align}

Now we take this upper bound further, specializing to the case of $d'=1$ and $N = \min(L_{\vp},M(d))$:
\begin{align*}
       \comp(\varepsilon,d)
       & \le T(\varepsilon^{-1},1, \vp) \Biggl [ \frac{\min(L_{\vp},M(d))}{T(\varepsilon^{-1},1, \vp)}  + \underbrace{\sum_{i = \min(L_{\vp},M(d))}^{M(d)} \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}}_{\le S_{\vp} + 1/ T(\lambda_{M(d), d}^{-1},1,\vp) \text{ by \eqref{eq:strong_rest_tractiff}} }
        \Biggr] \\
       & \le T(\varepsilon^{-1},1,\vp) \underbrace{\left[\frac{L_{\vp}+1}{T(0,1, \vp)} + S_{\vp} \right]}_{=:C_{\vp}}
       \qquad \text{by \eqref{eq:Tnondecreasing}}\\
       & =  C_{\vp}T(\varepsilon^{-1},1,\vp).
\end{align*}
This means that we have strong $T$-tractability on the restricted domain $\Omega$ via \eqref{eq:rest_strong_tract_def}, and verifies the sufficiency of \eqref{eq:strong_rest_tractiff}.

\bigskip
\noindent \textbf{Necessary condition:} \\
Suppose that we have strong
$T$-tractability on the restricted domain $\Omega$ as defined in \eqref{eq:rest_strong_tract_def}. That is, for some $\vp \ge \vzero$ there exists a positive constant $C_{\vp}$ such that
\[
\comp(\varepsilon,d)\le C_{\vp}\, T(\varepsilon^{-1},1,\vp)
\qquad \forall (\varepsilon^{-1}, d) \in \Omega.
\]
Since the sequence of eigenvalues $\lambda_{1,d}, \lambda_{2,d}, \ldots $ is non-increasing, we have
\begin{equation}\label{eq:lambda_K_strong4}
\lambda_{\lfloor C_{\vp}\, T(\varepsilon^{-1},1,\vp)\rfloor +1,d}\le \varepsilon \qquad \forall \varepsilon > E(d), \ d \in \mathbb{N}.
\end{equation}

For $(\varepsilon^{-1}, d) \in \Omega$, define
\[
i(\varepsilon,\vp):= \lfloor C_{\vp}\, T(\varepsilon^{-1},1,\vp)\rfloor +1, \quad
i(\infty,\vp) = \lim_{\varepsilon \to \infty}  i(\varepsilon,\vp) \ge 1 . 
\]
Thus, it follows by \eqref{eq:lambda_K_strong3} that $\lambda_{i(\varepsilon,\vp),d} \le \varepsilon$.
Note furthermore that we always have
\[
i(\varepsilon,\vp)\le C_{\vp}\, T(\varepsilon^{-1},1,\vp)+1 \le C_{\vp} T(\lambda_{i(\varepsilon,\vp),d}^{-1},1,\vp)+1, \qquad  \forall \varepsilon > E(d), \ d \in \mathbb{N},
\]
since
$T(\cdot,1, \vp)$ is non-decreasing.

For $\varepsilon$ taking on all values in $(E(d),\infty)$, $i(\varepsilon,\vp)$ takes on (at least) all values 
\[
I_{d,\vp}:=\begin{cases} 
\emptyset, & i(E(d),\vp) = i(\infty,\vp) \\
\{i(\infty,\vp),\ldots, i(E(d),\vp)-1\}, & E(d) > 0, \\
\{i(\infty,\vp), i(\infty,\vp) + 1, \ldots\}, & E(d) = 0.
\end{cases}
\]
So,
\[
i\le  C_{\vp} T(\lambda_{i,d}^{-1},1,\vp)+1, \qquad \forall i\in I_{d,\vp},  \ d \in \naturals.
\]
This implies via our technical assumption \eqref{eq:ptauassume} that
\begin{gather}
\nonumber
 K_{\vp,\tau}\,T (\lambda_{i,d}^{-1},1,\tau \vp) \ge
 [T(\lambda_{i,d}^{-1},1, \vp)]^\tau
 \ge
  \left[\frac{(i-1)}{C_{\vp}}\right]^\tau \qquad i\in I_{d,\vp},  \ d \in \naturals, \\
  \label{eq:rest_strong_ineq}
 \frac{1}{T (\lambda_{i,d}^{-1},1,\tau \vp)} \le
\frac{K_{\vp,\tau}\, C_{\vp}^\tau}{(i-1)^\tau}\qquad \forall i \in I_{d,\vp} \setminus\{1\}, \ d \in \naturals .
\end{gather}

To complete the proof, we will sum both sides of \eqref{eq:rest_strong_ineq} over the range of $i$ appearing in \eqref{eq:strong_rest_tractiff}, to establish that $S_{\tau\vp}$ is finite.  To do this we need to show that \eqref{eq:rest_strong_ineq} holds for that range.  Choose Choose $L_{\tau\vp} = i(\infty,\vp) +1 \ge 2$. There are three cases.
\begin{enumerate}
\renewcommand{\labelenumi}{\roman{enumi})}

\item If $M(d) < L_{\tau\vp}$, then 
\[
\sum_{i = L_{\tau\vp}}^{M(d)} \frac{1}{T (\lambda_{i,d}^{-1},1, \tau \vp)} = 0,
\]
which does not affect the finiteness of $S_{\tau\vp}$.

\item If $M(d) \ge L_{\tau\vp}$ and $E(d) = 0$, then $\{L_{\tau\vp},  \L_{\tau\vp} + 1, \ldots \} \subseteq I_{d,\vp} \setminus \{1\}$.

\item If $M(d) \ge L_{\tau\vp}$ and $E(d) >0$, then \eqref{eq:Mddef} and \eqref{eq:lambda_K_strong4} imply that
\[
\lambda_{i(E(d),\vp),d} \le \lambda_{M(d)+1,d} \le E(d) < \lambda_{M(d),d}.
\]
This means that $M(d) \le i(E(d),\vp)-1$.  So, $\{L_{\tau\vp}, \ldots, M(d)\} \subseteq I_{d,\vp} \setminus \{1\}$.

In both the second and third cases,
\begin{align*}
S_{\tau\vp} & = \sup_{d\in\naturals} \sum_{i = L_{\tau\vp}}^{M(d)} \frac{1}{T (\lambda_{i,d}^{-1},1, \tau \vp)}
\le 
\sup_{d\in\naturals} \sum_{i \in I_{d,\vp} \setminus\{1\} } \frac{1}{T (\lambda_{i,d}^{-1},1, \tau \vp)} \\
&  \le  K_{\vp,\tau}\, C_{\vp}^\tau
\sum_{i=2}^\infty \frac{1}{(i-1)^\tau}
 \le K_{\vp,\tau}\, C_{\vp}^\tau
\zeta (\tau) 
 < \infty \qquad \text{by \eqref{eq:rest_strong_ineq},}
\end{align*}
where $\zeta$ denotes the Riemann zeta function.

\end{enumerate}
Thus, in all cases, the assumption of strong tractability on the restricted domain implies \eqref{eq:strong_rest_tractiff} with $\vp$ replaced by $\vp' = \tau \vp$, and so we see the necessity of $\eqref{eq:strong_rest_tractiff}$.

\bigskip
\noindent \textbf{Optimality:} 
The proof of optimality is analogous to that for Theorem \ref{thm_main_strong_tract2} and is omitted.

@@

\noindent This concludes the proof of  Theorem \ref{thm_main_rest_strong_tract}.
\end{proof}

The equivalent condition for tractability on a restricted domain is analogous to the conditions for tractability on an unrestricted domain and for strong tractability on a restricted domain.  The proof is analogous as well.
